Verifying Spatial Properties of Array Computations

### 2017-04-17 oopsla17/study.tex

\noindent
At the beginning of this project, we made the following hypotheses
about array programming idioms:
\begin{enumerate}[leftmargin=2em]
\item Computations involving loops over arrays are common, where an
  arrays are read from with a static pattern based on constant offsets
  from (base or dervied) induction variables;

@@

\item Many array computations read from each particular index just once.
\end{enumerate}
%
From these hypotheses, we conjecture that the above programming idioms
can be specified declaratively via a small
set of combinators, capturing data access patterns based on
overlapping (hyper)rectangles. We performed a large scale source-code
analysis to validate these hypotheses and guide the design of our
language. These results are also of wider value to others, such as
designers of array/stencil DSLs or autotuners.

\subsection{Methodology}
%
\noindent
We constructed a corpus of \numPackages{} scientific computing
packages written in Fortran ranging in size and scope: The Unified
Model (UM)~\cite{um}, E3MG~\cite{RePEc:aen:journl:2006se-a12},
BLAS~\cite{blas}, Hybrid4~\cite{GBC:GBC635},
GEOS-Chem~\cite{geos-chem}, Navier (based on
\cite{griebel1997numerical}), Computational Physics 2 (CP2)
~\cite{nicholas2006computational}, ARPACK-NG~\cite{arpackng}, and
SPECFEM3D~\cite{specfem3d}. \dnote{Include the rest} This covers
approximately 1.4 million lines of physical Fortran code (2.4 million
including comments and white space).  Appendix~\ref{app:corpus}
provides further detail on these packages and their sizes, with UM and
GEOS-Chem the largest at $\approx$630kloc and $\approx$450kloc respectively. We used
Wheeler's \emph{SLOCCount} to get a count of the physical source lines
(excluding comments and blank lines)~\cite{wheeler2001sloccount}.

Half the packages come from active research teams who are our
project partners (\ie{}, they were not selected carefully to
fit our hypotheses). BLAS was selected as it is a common numerical library;
similarly CP2 was selected as it gives standard numerical
analysis algorithms from a popular computational physics textbook.
The smallest project Navier ($\approx$500loc) is based on a standard textbook on
computational fluid dynamics.

@@

\paragraph{Classifications}

We focus on assignment statements in loops where the right-hand side
expression reads from one or more arrays.  Left-hand sides and
right-hand sides are classified in the following way:

\begin{tabular}{l|l|l|l}
  Shorthand & Classification (of subscripts) & Applies to & Example \\ \hline
 (\textsf{vars}) & Variables & LHS & \fortran{x = ...}
  \\
 (\textsf{IVs}) & Induction variables & LHS & \fortran{a(i, j)} \\
 (\textsf{neigh}) & Neighbourhood offsets (of the form
                                                  $i \pm c$) & LHS/RHS
                                                               & \fortran{a(i, j-1)} \\
 (\textsf{neigh+c}) & Neighbourhood offsets and constants & LHS/RHS &
                                                            \fortran{b(i, 0, j+1)} \\
 (\textsf{aff}) & Affine offsets (of the form $a * i \pm b$) & LHS/RHS &
                                     \fortran{a(2*i+1,j)} \\
 (\textsf{aff+c}) & Affine offsets and constants & LHS/RHS & \fortran{a(i+1, 0,
                                                   3*j+2)} \\
 (\textsf{other}) & Subscript expression not included above & LHS/RHS &
\fortran{x(f(i))}, \fortran{a(i*i)}
\end{tabular} \\[1em]

\noindent
Note \textsf{vars} and \textsf{IVs} classifications apply only to
left-hand sides.
The classification is computed from all information that flows into an
assignment, for example, the following decomposes a one-dimensional
three-point pattern across multiple intermediate assignments:
%%
\begin{minted}{fortran}
do i = 1, n
  x = a(i)
  y = a(i+1)
  b(i) = (a(i-1) + x + y)/3.0
end do
\end{minted}
%%x
Our analysis recognises this as one array computation (rather than three), starting at
line 4, and reading from subscripts \fortran{a(i)}, \fortran{a(i+1)},
\fortran{a(i-1)}. We thus traverse the body of a loop in reverse,
using reaching-definitions to calculate of the array reads that flow
to a particular assignment. Since \fortran{x} and \fortran{y} flow to
the line 4, line 2 and line 3 are not classified, though multiple
assignments can flow to multiple classified statements.

We further sub-classify sets of array subscripts
on the right-hand side based on their spatial properties: \\

\begin{tabular}{l||l|p{0.28\linewidth}|l}
  Property   & Shorthand & Classifications (of RHS pattern) & Example \\ \hline
%%%%%%
\multirow{2}{*}{Contiguity} & (\textsf{contig}) & Contiguous  &
                                                                \fortran{a(i) + a(i+1) + a(i+2)} \\
  & (\textsf{disjoint}) & Non-contiguous & \fortran{a(i) + a(i+2)} \\ \hline
%%%%%%
\multirow{2}{*}{Shape} & (\textsf{rect}) & (Hyper)rectangle &
 \fortran{a(i,j) + a(i+1,j) + a(i,j+1) + a(i+1,j+1)} \\
           & (\textsf{sumRect}) & Composed (hyper)rectangles &
  \fortran{a(i,j) + a(i,j-1) + a(i-1,j) + a(i+1,j) + a(i,j+1)} \\ \hline
%%%%%%
  \multirow{2}{*}{Reuse} & (\textsf{readonce}) & Unique subscripts
                                         & \eg{} \fortran{b(i) = a(i) + a(i+1)} \\
  & (\textsf{mult}) & Repeated subscripts &
\eg{}  \fortran{b(i) = a(i)  + a(i)}
\end{tabular} \\[1em]

\noindent
To assess the hypothesis that
arrays are mostly read in a pattern that includes the immediate neighbours
to the ``origin'', we categorised the position of subscript patterns for each
right-hand side, in each dimension: \\

\begin{tabular}{l||l|p{0.33\linewidth}|l}
  Property   & Shorthand & Classification (indices per
                           dimension) & Example \\ \hline
%%%%%%
  \multirow{2}{*}{Positioning} & (\textsf{origin}) & Includes origin
& \fortran{a(i)} or \fortran{a(i+1, j)} (in $2^{\textit{nd}}$
  dimension) \\
             & (\textsf{straddle}) & Within distance 1 of origin &
\fortran{a(i+1), a(i-1)} \\
             & (\textsf{away}) & Away from the origin
                                          & \fortran{a(i+2), a(i+3)}
\end{tabular} \\[1em]

\noindent
Finally, we categorise the relationship between the left-hand side
and right-hand side, in terms of the use of induction variables.
The two sides are \emph{consistent} if the same induction variables
appear in each side, used for the same dimension. This is
weakened to a \emph{permutation} if the roles of the induction
variables changes. This is weakened further if the left-hand side
induction variables are either
a subset or superset of the induction variables on the right-hand
side. Otherwise, the two sides are seen as inconsistent:

\begin{tabular}{l|l|l}
  (\textsf{const}) & Consistent & \fortran{a(i, j) = b(i, j) +
                                  b(i+1,j+1)} \\
  (\textsf{perm}) & Permutation & \fortran{a(i, j) = c(j, i)} or \fortran{a(i,
                                                               0) =
                                 b(0, i)} \\
  (\textsf{LHSsub}) & LHS subset & \fortran{a(i) = b(i, j) + b(i, j-1)}
  \\
  (\textsf{LHSsup}) & LHS superset & \fortran{a(i, j) = b(i)} \\
  (\textsf{inconst}) & Inconsistent & \fortran{a(i) = b(j)}
\end{tabular} \\[1em]
%

@@

We revist each hypothesis and show the related data and conclusions.

@@

\item Loops over arrays mainly read from arrays with a static pattern
based on constant offsets from (base or dervied) induction variables;

@@

Neighbourhood + constants       & 23209     & 21.34\%  \\ \hline

@@

computations (88,577) have a static pattern that are either constants or
constant offsets of base/derived induction variables, confirming

@@

dimensions). The ``affine + constants'' class is

>>>>

\noindent
We start with the following hypotheses about array programming idioms
in scientific code, formed from our observations and conversations
with scientists:
\begin{enumerate}[leftmargin=2em]
\item Computations involving loops over arrays are common, where an
  arrays are read from with a static pattern based on constant offsets
  from (base or dervied) induction variables;

@@

\item Many array computations read from each particular array index just
  once. The significance of this hypothesis is that, if true, it would
  be useful to provide a specification of index uniqueness, so accidental
  repetition of subscripts in complex code can be detected.
\end{enumerate}
%
From these hypotheses, we conjecture that the spatial properties of
the above programming idioms can be specified declaratively via a
small set of combinators, capturing data access patterns.% based on
%overlapping (hyper)rectangles.
We performed a large scale source-code analysis to validate these
hypotheses and guide the design of our language. These results are
potentially of interest to others, such as designers of array/stencil DSLs
or autotuners.

\subsection{Methodology}
%
\noindent
We constructed a corpus of eleven scientific computing packages
written in Fortran ranging in size and scope:
%
\begin{enumerate}[leftmargin=1.5em]
\item The Unified Model (UM) developed by the UK Met Office
for weather modelling~\cite{um};
\item E3ME, a mixed economic/energy impact predication
  model~\cite{RePEc:aen:journl:2006se-a12}

\item Hybrid4, a global scale ecosystem model~\cite{GBC:GBC635};

\item GEOS-Chem, troposhperic chemstry model~\cite{geos-chem};

\item Navier, a small-size Navier-Stokes fluid model (based
  on \cite{griebel1997numerical};

\item Computational Physics (CP), programs from a popular
 textbook~\cite{giordano1997computational};

\item BLAS, a common linear-algebra
  library used in modelling~\cite{blas};

\item ARPACK-NG, an open-source library for solving large-scale
eigenvalue problems~\cite{arpackng};

\item SPECFEM3D, global seismic wave models~\cite{specfem3d};

\item MUDPACK, a general multi-grid solver for elliptical
partial differentials~\cite{MUD};

\item Cliffs, a tsunami model~\cite{tolkova2014land}
\end{enumerate}
%
These cover approximately 1.4 million lines of physical Fortran code
(2.4 million including comments and white space).  The UM and
GEOS-Chem packages are the largest at $\approx$630kloc and
$\approx$450kloc respectively.  Appendix~\ref{app:corpus} provides
further detail on these packages and their sizes.  We used Wheeler's
\emph{SLOCCount} to get a count of the physical source lines
(excluding comments and blank lines)~\cite{wheeler2001sloccount}.
A third of the packages come from active research teams who are our project
partners (\ie{}, they were not selected carefully to fit our
hypotheses).

@@

%\paragraph{Classifications}
Left-hand sides and right-hand sides of assignments are classified
based on their array subscripting expressions:
%
\begin{center}
\begin{tabular}{l|l|l}
 % Shorthand &
 Classification (of subscripts) & Applies to & Example \\ \hline
 %(\textsf{vars}) &
 Just a variable & LHS & \fortran{x = ...}
  \\
 %(\textsf{IVs}) &
 Induction variables & LHS & \fortran{a(i, j)} \\
 %(\textsf{neigh}) &
 Neighbourhood offsets (of the form
                                                  $i \pm c$) & LHS/RHS
                                                               & \fortran{a(i, j-1)} \\
 %(\textsf{neigh+c}) &
 Neighbourhood offsets and absolutes & LHS/RHS &
                                                            \fortran{b(i, 0, j+1)} \\
 %(\textsf{aff}) &
 Affine offsets (of the form $a * i \pm b$) & LHS/RHS &
                                     \fortran{a(2*i+1,j)} \\
 %(\textsf{aff+c}) &
 Affine offsets and absolutes & LHS/RHS & \fortran{a(i+1, 0,
                                                   3*j+2)} \\
 %(\textsf{other}) &
 Subscript expression not included above & LHS/RHS &
\fortran{x(f(i))}, \fortran{a(i*i)}, \fortran{a(0,1)}
\end{tabular}
\end{center}
%
%Note \textsf{vars} and \textsf{IVs} classifications apply only to
%left-hand sides.
Classifications are computed from all information that flows into an
assignment. For example, the following decomposes a one-dimensional
three-point pattern across multiple intermediate assignments:
%%
\begin{minted}{fortran}
do i = 1, n
  x = a(i)
  y = a(i+1)
  b(i) = (a(i-1) + x + y)/3.0
end do
\end{minted}
%%x
Our analysis recognises this as one array computation (rather than three), starting at
line 4, and reading from subscripts \fortran{a(i)}, \fortran{a(i+1)},
\fortran{a(i-1)}. We thus traverse the body of a loop in reverse,
using reaching-definitions to calculate the array reads that flow
to a particular assignment. Since \fortran{x} and \fortran{y} flow to
the line 4, line 2 and line 3 are not classified separately, though multiple
assignments can flow to multiple classified statements.

We further sub-classify sets of array subscripts
on the right-hand side based on their spatial properties:

\begin{center}
\begin{tabular}{l||l|p{0.28\linewidth}|l}
  Property   & Shorthand & Classifications (of RHS pattern) & Example \\ \hline
%%%%%%
\multirow{2}{*}{Contiguity} & (\textsf{contig}) & Contiguous  &
                                                                \fortran{a(i) + a(i+1) + a(i+2)} \\
  & (\textsf{disjoint}) & Non-contiguous & \fortran{a(i) + a(i+2)} \\ \hline
%%%%%%
%\multirow{2}{*}{Shape} & (\textsf{rect}) & (Hyper)rectangle &
% \fortran{a(i,j) + a(i+1,j) + a(i,j+1) + a(i+1,j+1)} \\
%           & (\textsf{sumRect}) & Composed (hyper)rectangles &
%  \fortran{a(i,j) + a(i,j-1) + a(i-1,j) + a(i+1,j) + a(i,j+1)} \\ \hline
%%%%%%
  \multirow{2}{*}{Reuse} & (\textsf{readonce}) & Unique subscripts
                                         & \fortran{b(i) = a(i) + a(i+1)} \\
  & (\textsf{mult}) & Repeated subscripts &
 \fortran{b(i) = a(i)  + a(i)} \\ \hline
  \multirow{2}{*}{Positioning} & (\textsf{origin}) & Includes origin
& \fortran{a(i)} or \fortran{a(i+1, j)} (in $2^{\textit{nd}}$
  dimension) \\
             & (\textsf{straddle}) & Within distance 1 of origin &
\fortran{a(i+1,j) + a(i-1,j)} \\
             & (\textsf{away}) & Away from the origin
                                          & \fortran{a(i+2), a(i+3)}
\end{tabular}
\end{center}
%
%To assess the hypothesis that
%arrays are mostly read in a pattern that includes the immediate neighbours
%to the ``origin'', we categorised the position of subscript patterns for each
%right-hand side, in each dimension: \\
%
%\noindent
%
Finally, we categorise the relationship between the left-hand side
and right-hand side, in terms of the use of induction variables.
The two sides are \emph{consistent} if the same induction variables
appear in each side, used for the same dimension. This is
weakened to a \emph{permutation} if the roles of the induction
variables changes. This is weakened further if the left-hand side
induction variables are either
a subset or superset of the induction variables on the right-hand
side. Otherwise, the two sides are seen as inconsistent:
%
\begin{center}
\begin{tabular}{l|l|l}
  Shorthand & Classification & Example \\ \hline
  (\textsf{const}) & Consistent & \fortran{a(i, j) = b(i, j) +
                                  b(i+1,j+1)} \\
  (\textsf{perm}) & Permutation & \fortran{a(i, j) = c(j, i)} or \fortran{a(i,
                                                               0) =
                                 b(0, i)} \\
  (\textsf{LHSsub}) & LHS subset & \fortran{a(i) = b(i, j) + b(i, j-1)}
  \\
  (\textsf{LHSsup}) & LHS superset & \fortran{a(i, j) = b(i)} \\
  (\textsf{inconst}) & Inconsistent & \fortran{a(i) = b(j)}
\end{tabular}
\end{center}
%

@@

We revisit each hypothesis and show the related data and conclusions.

@@

\item Computations involving loops over arrays are common, where an
  arrays are read from with a static pattern based on constant offsets
  from (base or dervied) induction variables;

@@

Neighbourhood + absolute       & 23209     & 21.34\%  \\ \hline

@@

computations (88,577) have a static pattern that comprises
constant offsets of base/derived induction variables (potential with
some absolute indices), confirming

@@

dimensions). The ``affine + absolute'' class is
